---
title: "DATA 606 - Lab 4a"
author: "Joshua Sturm"
date: "10/15/2017"
output:
  html_document:
    code_folding: show
    css: ./lab.css
    highlight: pygments
    smooth_scroll: yes
    theme: cerulean
    toc: yes
    toc_float: yes
    smart: no
  pdf_document:
    keep_tex: yes
always_allow_html: yes
---

In this lab, we investigate the ways in which the statistics from a random 
sample of data can serve as point estimates for population parameters.  We're 
interested in formulating a *sampling distribution* of our estimate in order 
to learn about the properties of the estimate, such as its distribution.

## The data

We consider real estate data from the city of Ames, Iowa.  The details of 
every real estate transaction in Ames is recorded by the City Assessor's 
office.  Our particular focus for this lab will be all residential home sales 
in Ames between 2006 and 2010.  This collection represents our population of 
interest.  In this lab we would like to learn about these home sales by taking 
smaller samples from the full population.  Let's load the data.

```{r load-data, eval=TRUE}
load("more/ames.RData")
```

We see that there are quite a few variables in the data set, enough to do a 
very in-depth analysis.  For this lab, we'll restrict our attention to just 
two of the variables: the above ground living area of the house in square feet 
(`Gr.Liv.Area`) and the sale price (`SalePrice`).  To save some effort 
throughout the lab, create two variables with short names that represent these 
two variables.  

```{r assign, eval=TRUE}
area <- ames$Gr.Liv.Area
price <- ames$SalePrice
```

Let's look at the distribution of area in our population of home sales by 
calculating a few summary statistics and making a histogram.

```{r area, eval=TRUE}
summary(area)
hist(area)
```

### Question 1
1.  Describe this population distribution.

#### Solution
This is a unimodal distribution, with a right skew, and a mean of ~1500.


## The unknown sampling distribution

In this lab we have access to the entire population, but this is rarely the 
case in real life.  Gathering information on an entire population is often 
extremely costly or impossible.  Because of this, we often take a sample of 
the population and use that to understand the properties of the population.

If we were interested in estimating the mean living area in Ames based on a 
sample, we can use the following command to survey the population.

```{r samp1, eval=TRUE}
# set seed to have a consistent output
set.seed(50)
samp1 <- sample(area, 50)
```

This command collects a simple random sample of size 50 from the vector 
`area`, which is assigned to `samp1`.  This is like going into the City 
Assessor's database and pulling up the files on 50 random home sales.  Working 
with these 50 files would be considerably simpler than working with all 2930 
home sales.

### Question 2
2.  Describe the distribution of this sample. How does it compare to the 
    distribution of the population?
    
#### Solution
```{r}
summary(samp1)
hist(samp1)
```
Although this is just a random sample of the total population, we can see that it's fairly simiarly to the overall population. It's also unimodal, with a right skew, and a similar mean.

If we're interested in estimating the average living area in homes in Ames 
using the sample, our best single guess is the sample mean.

```{r mean-samp1, eval=TRUE}
mean(samp1)
```

Depending on which 50 homes you selected, your estimate could be a bit above 
or a bit below the true population mean of 1499.69 square feet.  In general, 
though, the sample mean turns out to be a pretty good estimate of the average 
living area, and we were able to get it by sampling less than 3\% of the 
population.

### Question 3
3.  Take a second sample, also of size 50, and call it `samp2`.  How does the 
    mean of `samp2` compare with the mean of `samp1`?  Suppose we took two 
    more samples, one of size 100 and one of size 1000. Which would you think 
    would provide a more accurate estimate of the population mean?

#### Solution
```{r}
set.seed(805)
samp2 <- sample(area,50)
summary(samp2)
hist(samp2)
```

Once more, we can see that the distribution is similar to the overall population. 
Let's try with 100 and 1000:

```{r}
set.seed(549)
samp2_100 <- sample(area,100)
samp2_1000 <- sample(area,1000)
summary(samp2_100)
summary(samp2_1000)
```
It appears that the larger the sample size, the closer it is to the overall population.

Not surprisingly, every time we take another random sample, we get a different 
sample mean.  It's useful to get a sense of just how much variability we 
should expect when estimating the population mean this way. The distribution 
of sample means, called the *sampling distribution*, can help us understand 
this variability. In this lab, because we have access to the population, we 
can build up the sampling distribution for the sample mean by repeating the 
above steps many times. Here we will generate 5000 samples and compute the 
sample mean of each.

```{r loop, eval=TRUE}
sample_means50 <- rep(NA, 5000)

for(i in 1:5000){
   samp <- sample(area, 50)
   sample_means50[i] <- mean(samp)
   }

hist(sample_means50)
```

If you would like to adjust the bin width of your histogram to show a little 
more detail, you can do so by changing the `breaks` argument.

```{r hist-breaks, eval=TRUE}
hist(sample_means50, breaks = 25)
```

Here we use R to take 5000 samples of size 50 from the population, calculate 
the mean of each sample, and store each result in a vector called 
`sample_means50`. On the next page, we'll review how this set of code works.

### Question 4
4.  How many elements are there in `sample_means50`?  Describe the sampling 
    distribution, and be sure to specifically note its center.  Would you 
    expect the distribution to change if we instead collected 50,000 sample 
    means?

#### Solution
There are 5000 samples of 50. It appears to be normally distributed, with a mean of ~1500, which would match the overall population.

```{r}
set.seed(305)
sample_means500 <- rep(NA, 50000)

for(i in 1:50000){
   samp <- sample(area, 50)
   sample_means500[i] <- mean(samp)
   }

hist(sample_means500, breaks = 25)
```
Since the population is normally distributed, it hasn't changed much.

### Question 5
5.  To make sure you understand what you've done in this loop, try running a 
    smaller version.  Initialize a vector of 100 zeros called 
    `sample_means_small`.  Run a loop that takes a sample of size 50 from 
    `area` and stores the sample mean in `sample_means_small`, but only 
    iterate from 1 to 100.  Print the output to your screen (type 
    `sample_means_small` into the console and press enter).  How many elements 
    are there in this object called `sample_means_small`? What does each 
    element represent?
    
#### Solution
```{r}
sample_means_small <- rep(NA, 100)

for(i in 1:100)
{
  samp <- sample(area,50)
  sample_means_small[i] <- mean(samp) 
}
length(sample_means_small)
```
There are 100 elements, each containing a sample of size 50.

## Sample size and the sampling distribution

The sampling distribution that we computed tells us much about estimating 
the average living area in homes in Ames.  Because the sample mean is an 
unbiased estimator, the sampling distribution is centered at the true average 
living area of the the population, and the spread of the distribution 
indicates how much variability is induced by sampling only 50 home sales.

To get a sense of the effect that sample size has on our distribution, let's 
build up two more sampling distributions: one based on a sample size of 10 and 
another based on a sample size of 100.

```{r samp-10-100, eval=TRUE}
sample_means10 <- rep(NA, 5000)
sample_means100 <- rep(NA, 5000)

for(i in 1:5000){
  samp <- sample(area, 10)
  sample_means10[i] <- mean(samp)
  samp <- sample(area, 100)
  sample_means100[i] <- mean(samp)
}
```

To see the effect that different sample sizes have on the sampling 
distribution, plot the three distributions on top of one another.

```{r plot-samps, eval=TRUE, tidy = FALSE}
par(mfrow = c(3, 1))

xlimits <- range(sample_means10)

hist(sample_means10, breaks = 20, xlim = xlimits)
hist(sample_means50, breaks = 20, xlim = xlimits)
hist(sample_means100, breaks = 20, xlim = xlimits)
```

### Question 6
6.  When the sample size is larger, what happens to the center?  What about the spread?

#### Solution
When the sample size is larger, the distribution is more compressed, and more closely resembles a normal distribution.

* * *
## On your own

So far, we have only focused on estimating the mean living area in homes in 
Ames.  Now you'll try to estimate the mean home price.

### Question 1
Take a random sample of size 50 from `price`. Using this sample, what is your best point estimate of the population mean?

#### Solution
```{r}
# We can take the mean of the samples to estimate the mean of the population
sample_price_50 <- sample(price,50)
mean(sample_price_50)
```

### Question 2
Since you have access to the population, simulate the sampling distribution for $\bar{x}_{price}$ by taking 5000 samples from the population of size 50 and computing 5000 sample means.  Store these means in a vector called `sample_means50`. Plot the data, then describe the shape of this sampling distribution. Based on this sampling distribution, what would you guess the mean home price of the population to be? Finally, calculate and report the population mean.

#### Solution
```{r}
sample_means50 <- rep(NA,5000)

for(i in 1:5000)
{
  samp <- sample(price,50)
  sample_means50[i] <- mean(samp)
}
hist(sample_means50)
mean(sample_means50)
mean(price)
```
It appears to be normally distributed. Furthermore, the mean of the sample data is very close to the mean of the population.

### Question 3
Change your sample size from 50 to 150, then compute the sampling distribution using the same method as above, and store these means in a new vector called `sample_means150`. Describe the shape of this sampling distribution, and compare it to the sampling distribution for a sample size of 50.  Based on this sampling distribution, what would you guess to be the mean sale price of homes in Ames?

#### Solution
```{r}
sample_means150 <- rep(NA,5000)

for(i in 1:5000)
{
  samp <- sample(price,150)
  sample_means150[i] <- mean(samp)
}
par(mfrow = c(2,1))
hist(sample_means50)
hist(sample_means150)
mean(sample_means150)
mean(price)
```
This sample is also normally distributed, similar to that of the previous sample. We can see that the mean of this sample is even closer to the mean for the population.

### Question 4
Of the sampling distributions from 2 and 3, which has a smaller spread?  If we're concerned with making estimates that are more often close to the true value, would we prefer a distribution with a large or small spread?

#### Solution
As we saw just before question 6, the larger the sample size, the smaller the spread is. In this case, problem 3 (sample size 150) has a smaller spread. In general, a smaller spread is preferred, since it's more representative of the overall population.

