---
title: "DATA 606 - Lab 8"
author: "Joshua Sturm"
output:
  pdf_document:
    keep_tex: yes
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    code_folding: show
    toc: yes
    toc_float: yes
    smart: no
always_allow_html: yes
---

## Grading the professor

Many college courses conclude by giving students the opportunity to evaluate 
the course and the instructor anonymously. However, the use of these student 
evaluations as an indicator of course quality and teaching effectiveness is 
often criticized because these measures may reflect the influence of 
non-teaching related characteristics, such as the physical appearance of the 
instructor. The article titled, "Beauty in the classroom: instructors' 
pulchritude and putative pedagogical productivity" (Hamermesh and Parker, 2005) 
found that instructors who are viewed to be better looking receive higher 
instructional ratings. (Daniel S. Hamermesh, Amy Parker, Beauty in the 
classroom: instructors pulchritude and  putative pedagogical productivity, 
*Economics of Education Review*, Volume 24, Issue 4, August 2005, Pages 369-376, 
ISSN 0272-7757, 10.1016/j.econedurev.2004.07.013. [http://www.sciencedirect.com/science/article/pii/S0272775704001165](http://www.sciencedirect.com/science/article/pii/S0272775704001165).)

In this lab we will analyze the data from this study in order to learn what goes 
into a positive professor evaluation.

## The data

The data were gathered from end of semester student evaluations for a large 
sample of professors from the University of Texas at Austin. In addition, six 
students rated the professors' physical appearance. (This is aslightly modified 
version of the original data set that was released as part of the replication 
data for *Data Analysis Using Regression and Multilevel/Hierarchical Models* 
(Gelman and Hill, 2007).) The result is a data frame where each row contains a 
different course and columns represent variables about the courses and professors.

```{r load-data, eval=TRUE}
load("more/evals.RData")
```

variable         | description
---------------- | -----------
`score`          | average professor evaluation score: (1) very unsatisfactory - (5) excellent.
`rank`           | rank of professor: teaching, tenure track, tenured.
`ethnicity`      | ethnicity of professor: not minority, minority.
`gender`         | gender of professor: female, male.
`language`       | language of school where professor received education: english or non-english.
`age`            | age of professor.
`cls_perc_eval`  | percent of students in class who completed evaluation.
`cls_did_eval`   | number of students in class who completed evaluation.
`cls_students`   | total number of students in class.
`cls_level`      | class level: lower, upper.
`cls_profs`      | number of professors teaching sections in course in sample: single, multiple.
`cls_credits`    | number of credits of class: one credit (lab, PE, etc.), multi credit.
`bty_f1lower`    | beauty rating of professor from lower level female: (1) lowest - (10) highest.
`bty_f1upper`    | beauty rating of professor from upper level female: (1) lowest - (10) highest.
`bty_f2upper`    | beauty rating of professor from second upper level female: (1) lowest - (10) highest.
`bty_m1lower`    | beauty rating of professor from lower level male: (1) lowest - (10) highest.
`bty_m1upper`    | beauty rating of professor from upper level male: (1) lowest - (10) highest.
`bty_m2upper`    | beauty rating of professor from second upper level male: (1) lowest - (10) highest.
`bty_avg`        | average beauty rating of professor.
`pic_outfit`     | outfit of professor in picture: not formal, formal.
`pic_color`      | color of professor's picture: color, black & white.

## Exploring the data

### Question 1
Is this an observational study or an experiment? The original research question posed in the paper is whether beauty leads directly to the differences in course evaluations. Given the study design, is it possible to answer this question as it is phrased? If not, rephrase the question.

#### Solution
Since there is no control group, this is an observational study. We cannot determine causation from an observational study, so a better quesiton would be: "Is there a correlation between the professor's appearance, and their evaluated score?"
  

### Question 2
Describe the distribution of `score`. Is the distribution skewed? What does that tell you about how students rate courses? Is this what you expected to see? Why, or why not?

#### Solution
```{r quest-2}
hist(evals$score)
```
The distribution is unimodal, with a left skew, which means students tend to give higher ratings. This is surprising to me, since research has shown that people are more likely to leave a review when they are dissatisfied.

### Question 3
Excluding `score`, select two other variables and describe their relationship using an appropriate visualization (scatterplot, side-by-side boxplots, or mosaic plot).

#### Solution
```{r quest-3}
plot(evals$cls_perc_eval ~ evals$gender)
```

The percentage of the class that fills out the evaluation doesn't differ too much based on the gender of the professor. 

## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking teachers
are evaluated more favorably. Let's create a scatterplot to see if this appears 
to be the case:

```{r scatter-score-bty_avg}
plot(evals$score ~ evals$bty_avg)
```


Before we draw conclusions about the trend, compare the number of observations 
in the data frame with the approximate number of points on the scatterplot. 
Is anything awry?

There were 463 observations, but the scatterplot has fewer points.

### Question 4
Replot the scatterplot, but this time use the function `jitter()` on the $y$- or the $x$-coordinate. (Use `?jitter` to learn more.) What was misleading about the initial scatterplot?

#### Solution
```{r quest-4}
plot(jitter(evals$score)~evals$bty_avg)
```

We can see that there are many points on top of each other, which is why the scatterplot had fewer points.

### Question 5
Let's see if the apparent trend in the plot is something more than natural variation. Fit a linear model called `m_bty` to predict average professor score by average beauty rating and add the line to your plot using `abline(m_bty)`. Write out the equation for the linear model and interpret the slope. Is average beauty score a statistically significant predictor? Does it appear to be a practically significant predictor?

#### Solution
```{r quest-5}
m_bty <- lm(score ~ bty_avg, data = evals)
plot(jitter(evals$score) ~ evals$bty_avg)
abline(m_bty)
summary(m_bty)
```
$\widehat{score} =$ `r m_bty$coefficients[[1]]` + `r m_bty$coefficients[[2]]` $\times \text{bty_avg}$

Average beauty score is significant, since its p-value is nearly 0.
It's not a practically significant predictor, though, since every additional point in `bty_avg` results in `r m_bty$coefficients[[2]]` additional points in overall score.

### Question 6
Use residual plots to evaluate whether the conditions of least squares regression are reasonable. Provide plots and comments for each one (see the Simple Regression Lab for a reminder of how to make these).

#### Solution
```{r quest-6}
plot(m_bty$residuals ~ evals$bty_avg)
abline(h = 0, lty = 3)
hist(m_bty$residuals)
qqnorm(m_bty$residuals)
qqline(m_bty$residuals)
```

The histogram is left-skewed, and the plot does not follow the diagonal line. Since the conditions for least squares regression are not met, we cannot use it in this case.

## Multiple linear regression

The data set contains several variables on the beauty score of the professor: 
individual ratings from each of the six students who were asked to score the 
physical appearance of the professors and the average of these six scores. Let's 
take a look at the relationship between one of these scores and the average 
beauty score.

```{r bty-rel}
plot(evals$bty_avg ~ evals$bty_f1lower)
cor(evals$bty_avg, evals$bty_f1lower)
```

As expected the relationship is quite strong - after all, the average score is 
calculated using the individual scores. We can actually take a look at the 
relationships between all beauty variables (columns 13 through 19) using the 
following command:

```{r bty-rels}
plot(evals[,13:19])
```

These variables are collinear (correlated), and adding more than one of these 
variables to the model would not add much value to the model. In this 
application and with these highly-correlated predictors, it is reasonable to use
the average beauty score as the single representative of these variables.

In order to see if beauty is still a significant predictor of professor score 
after we've accounted for the gender of the professor, we can add the gender 
term into the model.

```{r scatter-score-bty_avg_gender}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
summary(m_bty_gen)
```

### Question 7
P-values and parameter estimates should only be trusted if the conditions for the regression are reasonable. Verify that the conditions for this model are reasonable using diagnostic plots.

#### Solution
```{r quest-7}
plot(m_bty_gen$residuals ~ evals$bty_avg)
abline(h = 0, lty = 3)

hist(m_bty_gen$residuals)

qqnorm(m_bty_gen$residuals)
qqline(m_bty_gen$residuals)
```
The histogram is left-skewed, and the plot does not follow the diagonal line. The conditions for least square regression are not met, and it cannot be used in this case.

### Question 8
Is `bty_avg` still a significant predictor of `score`? Has the addition of `gender` to the model changed the parameter estimate for `bty_avg`?

#### Solution
The addition of `gender` has reduced the p-value even further, making `bty_avg` an even more significant predictor.

Note that the estimate for `gender` is now called `gendermale`. You'll see this 
name change whenever you introduce a categorical variable. The reason is that R 
recodes `gender` from having the values of `female` and `male` to being an 
indicator variable called `gendermale` that takes a value of $0$ for females and
a value of $1$ for males. (Such variables are often referred to as "dummy" 
variables.)

As a result, for females, the parameter estimate is multiplied by zero, leaving 
the intercept and slope form familiar from simple regression.

\[
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (0) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg\end{aligned}
\]

We can plot this line and the line corresponding to males with the following 
custom function.

```{r twoLines}
multiLines(m_bty_gen)
```

### Question 9
What is the equation of the line corresponding to males? (*Hint:* For males, the parameter estimate is multiplied by 1.) For two professors who received the same beauty rating, which gender tends to have the higher course evaluation score?

#### Solution
```{r eval = T, echo = F}
sumco <- summary(m_bty_gen)$coefficients
b0 <- sumco[[1]]
b1 <- sumco[[2]]
b2 <- sumco[[3]]
```

$\widehat{score} = \hat{\beta_0} + \hat{\beta_1} \times \text{bty_avg} + \hat{\beta_2} \times \text{Male}$
$\widehat{score} = `r b0` + `r b1` \times \text{bty_avg} + `r b2` \times 1$
$\widehat{score} = `r b0 + b2` + `r b1` \times \text{bty_avg}$.

Of the two professors with the same beauty rating, the male's evaluation score will be higher by `r b2`.

The decision to call the indicator variable `gendermale` instead of`genderfemale`
has no deeper meaning. R simply codes the category that comes first 
alphabetically as a $0$. (You can change the reference level of a categorical 
variable, which is the level that is coded as a 0, using the`relevel` function. 
Use `?relevel` to learn more.)

### Question 10
Create a new model called `m_bty_rank` with `gender` removed and `rank` added in. How does R appear to handle categorical variables that have more than two levels? Note that the rank variable has three levels: `teaching`, `tenure track`, `tenured`.

#### Solution
```{r quest-10}
m_bty_rank <- lm(score ~ bty_avg + rank, data = evals)
summary(m_bty_rank)
multiLines(m_bty_rank)
```
R creates dummy variables for each level.

The interpretation of the coefficients in multiple regression is slightly 
different from that of simple regression. The estimate for `bty_avg` reflects
how much higher a group of professors is expected to score if they have a beauty
rating that is one point higher *while holding all other variables constant*. In
this case, that translates into considering only professors of the same rank 
with `bty_avg` scores that are one point apart.

## The search for the best model

We will start with a full model that predicts professor score based on rank, 
ethnicity, gender, language of the university where they got their degree, age, 
proportion of students that filled out evaluations, class size, course level, 
number of professors, number of credits, average beauty rating, outfit, and 
picture color.

### Question 11
Which variable would you expect to have the highest p-value in this model? Why? *Hint:* Think about which variable would you expect to not have any association with the professor score.
  
#### Solution
I would guess `language` would have the least impact.

Let's run the model...

```{r m_full, eval = TRUE, tidy = FALSE}
m_full <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

### Question 12
Check your suspicions from the previous exercise. Include the model output in your response.

#### Solution
The variable with the highest p-value is `cls_profs` with 0.77806, compared to 0.03965 for `language`. I guessed wrong.

### Question 13
Interpret the coefficient associated with the ethnicity variable.

#### Solution
Keeping all other variables constant, professors that are not a minority will have a score higher by 0.1234929.

### Question 14
Drop the variable with the highest p-value and re-fit the model. Did the coefficients and significance of the other explanatory variables change? (One of the things that makes multiple regression interesting is that coefficient estimates depend on the other variables that are included in the model.) If not, what does this say about whether or not the dropped variable was collinear with the other explanatory variables?

#### Solution
```{r quest-14}
m_full_refit <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full_refit)
```

Yes, the coefficients and significance of the other variables did change, but not much, implying there wasn't much dependence between the other variables and `cls_profs`.

### Question 15
Using backward-selection and p-value as the selection criterion, determine the best model. You do not need to show all steps in your answer, just the output for the final model. Also, write out the linear model for predicting score based on the final model you settle on.

#### Solution
```{r quest-15}
m_full_best <- lm(score ~ ethnicity + gender + language + age + cls_perc_eval 
             + cls_credits + bty_avg + pic_color, data = evals)
summary(m_full_best)
sumco.2 <- summary(m_full_best)$coefficients
```
Removed all variables with a p-value higher than 0.05.

$$ \widehat{score} = `r sumco.2[1]` + `r sumco.2[2]` \times \text{ethnicity} + `r sumco.2[3]` \times \text{gender}  \\
`r sumco.2[4]` \times \text{language}`r sumco.2[5]` \times \text{age} + `r sumco.2[6]` \times \text{cls_perc_eval} \\
+ `r sumco.2[7]` \times \text{cls_credits} + `r sumco.2[8]` \times \text{bty_avg} `r sumco.2[9]` \times \text{pic_color}$$

### Question 16
Verify that the conditions for this model are reasonable using diagnostic plots.

#### Solution
```{r quest-16}
plot(m_full_best$residuals ~ evals$bty_avg)
abline(h = 0, lty = 3)

hist(m_full_best$residuals)

qqnorm(m_full_best$residuals)
qqline(m_full_best$residuals)
```
The histogram is left-skewed and the plot does not follow the diagonal line. The residuals do not appear to be nearly normal.

### Question 17
The original paper describes how these data were gathered by taking a sample of professors from the University of Texas at Austin and including all courses that they have taught. Considering that each row represents a course, could this new information have an impact on any of the conditions of linear regression?

#### Solution
It's possible that some students had the same professor more than once, which may affect the independence condition of the model.

### Question 18
Based on your final model, describe the characteristics of a professor and course at University of Texas at Austin that would be associated with a high evaluation score.

#### Solution
The professors with the highest scores would be male, not a minority, younger, educated in an english-speaking university, majority of their students completed the evaluation, teach single-credit courses, above-average looks, and have a black-and-white picture. 

### Question 19
Would you be comfortable generalizing your conclusions to apply to professors generally (at any university)? Why or why not?

#### Solution
No. This was an observational study performed on professors in a single university. Different universities, with different cultures, size, ranking, etc, may fare differently.