---
title: "DATA 606 - Homework 4"
author: "Joshua Sturm"
date: "10/15/2017"
output:
  html_document:
    code_folding: show
    smooth_scroll: yes
    toc: yes
    toc_float: yes
    smart: no
  pdf_document:
    keep_tex: yes
    toc: yes
always_allow_html: yes
---

## 4.4 Heights of adults
Researchers studying anthropometry collected body girth measurements and skeletal diameter measurements, as well as age, weight, height and gender, for 507 physically active individuals. The histogram below shows the sample distribution of heights in centimeters.

### (a) 
What is the point estimate for the average height of active individuals? What about the median? 

#### Solution
The point estimate for the average is the mean: 171.1. Similarly, the point estimate for the median would be 170.3.

### (b)
What is the point estimate for the standard deviation of the heights of active individuals? What about the IQR?

#### Solution
The point estimate for the standard deviation is 9.4. 
The IQR is Q3 - Q1: 177.8 - 163.8 = 14.

### (c)
Is a person who is 1m 80cm (180 cm) tall considered unusually tall? And is a person who is 1m 55cm (155cm) considered unusually short? Explain your reasoning.

#### Solution
They are both unusual in that they are above the third quartile and below the first quartile, respectively.

```{r}
n <- 507
mean <- 171.1
sd <- 9.4
shorter <- 155
taller <- 180
# Calculate z-scores:
z_shorter <- (shorter - mean) / sd
z_taller <- (taller - mean) / sd
z_shorter
z_taller
```
Looking at the z-scores, though, the person with a height of 155cm is more unusual, since they are considerably farther from the mean.

### (d)
The researchers take another random sample of physically active individuals. Would you expect the mean and the standard deviation of this new sample to be the ones given above? Explain your reasoning.

#### Solution
I would expect the mean and sd to be similar but different from the first sample.

### (e)
The sample means obtained are point estimates for the mean height of all active individuals, if the sample of individuals is equivalent to a simple random sample. What measure do we use to quantify the variability of such an estimate (Hint: recall that $SD_{\bar x} = \frac{\sigma}{\sqrt{n}}$)? Compute this quantity using the data from the original sample under the condition that the data are a simple random sample.

#### Solution
We can use the Standard Error: $SE = \frac{\sigma}{\sqrt{n}}$.
```{r}
SE <- sd / sqrt(n)
SE
```

## 4.14 Thanksgiving spending, Part I. 
The 2009 holiday retail season, which kicked off on November 27, 2009 (the day after Thanksgiving), had been marked by somewhat lower self-reported consumer spending than was seen during the comparable period in 2008. To get an estimate
of consumer spending, 436 randomly sampled American adults were surveyed. Daily consumer spending for the six-day period after Thanksgiving, spanning the Black Friday weekend and Cyber Monday, averaged \$84.71. A 95% confidence interval based on this sample is (\$80.31, $89.11). Determine whether the following statements are true or false, and explain your reasoning.

### (a)
We are 95% confident that the average spending of these 436 American adults is between \$80.31 and \$89.11.

#### Solution
**False.**
The average spent in this sample is \$84.71. We are 95% confident that the average for the entire *population* is between \$80.31 and \$89.11.

### (b)
This confidence interval is not valid since the distribution of spending in the sample is right skewed.

#### Solution
**False.**
So long as the sample is larger than 30, it's valid, and the skew should not matter.

### (c)
95% of random samples have a sample mean between \$80.31 and \$89.11.

#### Solution
**False.**
This confidence interval provides us with an idea of where the population mean is. A sample can differ from this, since the confidence interval is based on the mean and sd, which varies between samples.

### (d)
We are 95% confident that the average spending of all American adults is between \$80.31 and \$89.11.

#### Solution
**True.**
This is the idea behind the confidence interval - to find the population mean.

### (e)
A 90% confidence interval would be narrower than the 95% confidence interval since we don't need to be as sure about our estimate.

#### Solution
**True.**
If we're not measuring as precisely, the interval will be smaller, and less accurate.

### (f)
In order to decrease the margin of error of a 95% confidence interval to a third of what it is now, we would need to use a sample 3 times larger.

#### Solution
**False.**
The formula for the margin of error is: $MOE = z* \times SE$.
If we want $\frac{1}{3}\times SE$, we get $\frac{\sigma}{\sqrt(3^2 * n)}$, so we'd have to use a sample 9 times larger.

### (g)
The margin of error is 4.4.

#### Solution
**True.**
```{r}
confl <- 80.31
confu <- 89.11
moe <- (confu - confl) / 2
moe
```

## 4.24 Gifted children, Part I. 
Researchers investigating characteristics of gifted children collected data from schools in a large city on a random sample of thirty-six children who were identified as gifted children soon after they reached the age of four. The following histogram shows the distribution of the ages (in months) at which these children first counted to 10 successfully. Also provided are some sample statistics.

### (a)
Are conditions for inference satisfied?

#### Solution
I believe so.
Since the sample is from a large pool (school children in a large city), we can assume they're independent.
Furthermore, we have a sample size of 36, so $n = 36 \geq 30$.
Lastly, there doesn't appear to be any strong skew in the histogram plot.

### (b)
Suppose you read online that children first count to 10 successfully when they are 32 months old, on average. Perform a hypothesis test to evaluate if these data provide convincing evidence that the average age at which gifted children fist count to 10 successfully is less than the general average of 32 months. Use a significance level of 0.10.

#### Solution
Null hypothesis $H_0: \mu = 32$ months.\
Alternate hypothesis $H_A: \mu < 32$ months.\
$\alpha  = 0.10$.\
Since we're only interested if it's less than the average, it's a one-sided hypothesis test.
```{r}
n <- 36
mean <- 30.69
sd <- 4.31
SE <- sd / sqrt(n)
z <- ((mean - 32)/SE)
pnorm(z)
```

### (c)
Interpret the p-value in context of the hypothesis test and the data.

#### Solution
Because the p-value is less than the significance level, we reject the null hypothesis, and go with the alternate, i.e. gifted children can count to 10 in less than 32 months.

### (d) 
Calculate a 90% confidence interval for the average age at which gifted children first count to 10 successfully.

#### Solution
```{r}
confl <- mean - (1.65 * SE)
confu <- mean + (1.65 * SE)
cat(confl,confu)
```
We are 90% confident that the average age gifted children first count to 10 is between 29.50 and 31.88 months.

### (e)
Do your results from the hypothesis test and the confidence interval agree? Explain.

#### Solution
Yes. We rejected the null hypothesis of 32 months, and the confidence interval does not include 32.

## 4.26 Gifted children, Part II.
Exercise 4.24 describes a study on gifted children. In this study, along with variables on the children, the researchers also collected data on the mother's and father's IQ of the 36 randomly sampled gifted children. The histogram below shows the distribution of mother's IQ. Also provided are some sample statistics.

### (a)
Perform a hypothesis test to evaluate if these data provide convincing evidence that the average IQ of mothers of gifted children is different than the average IQ for the population at large, which is 100. Use a significance level of 0.10.

#### Solution
Null hypothesis $H_0: \mu = 100$ (IQ).\
Alternate hypothesis $H_A: \mu \neq 100$ (IQ).\
$\alpha  = 0.10$.\
Since we're interested if it differs from the average, it's a two-sided hypothesis test.
```{r}
n <- 36
mean <- 118.2
sd <- 6.5
SE <- sd/sqrt(n)
z <- ((mean - 100)/SE)
2*(1-pnorm(z)) # symmetric
```
Since the p-value is approximately 0, which is less than the significance level of 0.10, we can reject the null hypothesis in favour of the alternative.

### (b)
Calculate a 90% confidence interval for the average IQ of mothers of gifted children.

#### Solution
```{r}
confl <- mean - (1.65 * SE)
confu <- mean + (1.65 * SE)
cat(confl,confu)
```
We are 90% confident that the average IQ of mothers of gifted children is between 116.4 and 120.

### (c)
Do your results from the hypothesis test and the confidence interval agree? Explain.

#### Solution
Yes. We rejected the null hypothesis because the p-value was too low, and the given mean is not inside the confidence interval.

## 4.34 CLT. 
Define the term "sampling distribution" of the mean, and describe how the shape, center, and spread of the sampling distribution of the mean change as sample size increases.

### Solution
The sampling distribution is the distribution of the means of a collection of points estimates of a population. As the sample size increases, it more closely resembles the normal distribution, has a narrower spread, and the center is closer to the true population mean.

## 4.40 CFLBs.
A manufacturer of compact fluorescent light bulbs advertises that the distribution of the lifespans of these light bulbs is nearly normal with a mean of 9,000 hours and a standard deviation of 1,000 hours.

### (a)
What is the probability that a randomly chosen light bulb lasts more than 10,500 hours?

#### Solution
```{r}
n <- 10500
mean <- 9000
sd <- 1000
z = (n - mean)/sd

# since we want x > 10,500, we look for 1-pnorm(z)
1-pnorm(z)
```
The probability that a randomly chosen buld lasts more than 10,500 hours is ~6.68%.

### (b)
Describe the distribution of the mean lifespan of 15 light bulbs.

#### Solution
The sample size of 15 is small (<30), but we can assume that it will be nearly normal.
```{r}
SE <- sd/sqrt(15)
SE
```

### (c)
What is the probability that the mean lifespan of 15 randomly chosen light bulbs is more than 10,500 hours?

#### Solution
```{r}
z <- (n - mean)/SE
# Since we want x > 10,500, we look for 1-pnorm(z)
1-pnorm(z)
```
The probability that the mean lifespan of 15 random bulbs is > 10,500 is approximately 0.

### (d)
Sketch the two distributions (population and sampling) on the same scale.

#### Solution
```{r}
library(ggplot2)
seq <- seq(6000,12000)
dp <- dnorm(seq, mean, sd)
ds <- dnorm(seq, mean, SE)

ggplot() +
  geom_line(aes(seq, ds, color = "Population")) +
  geom_line(aes(seq,dp, color = "Sample")) +
  labs(x = "Hours", y = "Density")
```

#### (e)
Do your results from the hypothesis test and the confidence interval agree? Explain.

#### Solution
No. For (a), we'd need a nearly normal distribution to estimate with a sample. Similarly, we'd need a larger sample size ($> 30$) in (c) for it to be nearly normal.

## 4.48 Same observation, different sample size.
Suppose you conduct a hypothesis test based on a sample where the sample size is n = 50, and arrive at a p-value of 0.08. You then refer back to your notes and discover that you made a careless mistake, the sample size should have been n = 500. Will your p-value-increase, decrease, or stay the same? Explain.

### Solution
The formula for $Z$ is $Z = \frac{\overline{x} - \text{null value}}{SE_{\overline{x}}}$.\
$SE = \frac{\sigma}{\sqrt{n}}$, so as $n$ increases, $SE$ decreases. Consequently, we get a larger $Z$, and, therefore, a smaller $p$.











